### 2021-07-07 app_ch3.tex
this is a nightmare
progress ?

\chapter{Calculations involved in the bed elevation model}

\section{Numerical simulation algorithm}

The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
In the following sections, we begin by demonstrating the time intervals $\tau$ between subsequent transitions (i.e., the intervals between transition times) are exponentially distributed within the model we develop in the main text. Then we describe the SSA as a consequence of this property.

@@

To initialize a simulation, we specify the initial conditions $n_0$ and $m_0$, the model parameters for use in equations (2-7) in the main paper, and the desired simulation duration $t_\text{max}$. The SSA uses these inputs to generate time series of $n$ and $m$ as follows:

@@

By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, we obtain the mean field equation for the particle activity: 
\be 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1)+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\ee
This can be solved by introducing the generating function \citep{Cox1965} $G(x) = \sum_n x^n A(n)$, providing
\be 0 = (\nu+\lambda)(x-1)G + [\mu x^2 + \sigma + \gamma - (\mu + \sigma + \gamma)x]\frac{\partial G}{\partial x},\ee
which is separable and integrates for 
\be G(x) = \Bigg( \frac{\gamma + \sigma -\mu}{\gamma + \sigma - \mu x}\Bigg)^{\frac{\nu + \lambda}{\mu}}\ee
after applying the normalization condition $G(1)=1$.
From the definition of $G$ we can determine $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation (10) of the manuscript.

@@

From the negative binomial distribution we have $\langle n |m \rangle = \langle n \rangle$, so equation (9) provides 
\begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m). \end{multline}
Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation (11) of the manuscript
\be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
The \citet{Martin2014} Ornstein-Uhlenbeck model.

We solve this using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small. Introducing $z$ and writing $\overline{\kappa}=\kappa/z_1$ gives
\be 0 = [1+\overline{\kappa}(z+z_1)]M(z+z_1) + [1-\overline{\kappa}(z-z_1)]M(z-z_1) - 2M(z).\ee
$\overline{\kappa}$ should not depend on $z_1$ since the magnitude of the feedbacks between bed elevations and entrainment and deposition rates depends on elevation changes, and not on the size of grains or the length of the control volume.
Expanding the entire first and second terms to second order $z_1$ provides the Fokker-Planck equation
\be 0 = -2\overline{\kappa}z_1[zM(z)]' + z_1^2 M''(z). \ee
Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, we find
\be M = M_0e^{-\kappa (z/z_1)^2}\ee
as stated in the manuscript.

\subsection{Closure equation approach for bed elevations}
We set out to determine an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
\be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
for the mean particle activity conditional to the elevation $m$ into into manuscript equation (9), noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
\be 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
\ee
At this point, we consider $c$ an undetermined positive constant that may depend on the entrainment, migration, and deposition rates.
Taking the Fokker-Planck expansion and requiring the distribution $M(m)$ to vanish at infinity for normalization yields
\be 0 = 2\kappa m \Big( 2 + \frac{(\sigma-\mu)c}{E}\Big)M(m) + \Big\{
\Big[2-\kappa m \frac{(\sigma+\mu)c}{E} \Big]M(m)\Big\}'.\ee
Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$, we find
\be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}.\ee
Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, we determine the closure relation
\be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation (12) of the manuscript that is plotted with the numerical simulations in manuscript figure (4a).

>>>>

\chapter{Calculations involved in the bed elevation model}
\label{ch:gill}
\section{Numerical simulation algorithm}

The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of chapter \ref{ch:ch3}. The SSA follows as a consequence of this property.

@@

Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in equations \ref{eq:rate1}-\ref{eq:rate4}, and the desired simulation duration $t_\text{max}$. The SSA uses these inputs to generate time series of $n$ and $m$ as follows:

@@

By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, one can obtain the mean field equation for the particle activity: 
\be 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1)+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\ee
This can be solved by introducing the generating function \citep{Cox1965} $G(x) = \sum_n x^n A(n)$, providing
\be 0 = (\nu+\lambda)(x-1)G + [\mu x^2 + \sigma + \gamma - (\mu + \sigma + \gamma)x]\frac{\partial G}{\partial x},\ee
which is separable and integrates for 
\be G(x) = \Bigg( \frac{\gamma + \sigma -\mu}{\gamma + \sigma - \mu x}\Bigg)^{\frac{\nu + \lambda}{\mu}}\ee
after applying the normalization condition $G(1)=1$.
From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation \ref{eq:ancey} and introduced originally in section \ref{sec:birthdeath}.

@@

The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so equation (9) produces 
\begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m). \end{multline}
Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation (11) of the manuscript
\be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
The \citet{Martin2014} Ornstein-Uhlenbeck model.

This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small. Introducing $z$ and writing $\overline{\kappa}=\kappa/z_1$ gives
\be 0 = [1+\overline{\kappa}(z+z_1)]M(z+z_1) + [1-\overline{\kappa}(z-z_1)]M(z-z_1) - 2M(z).\ee
$\overline{\kappa}$ should not depend on $z_1$ since the magnitude of the feedbacks between bed elevations and entrainment and deposition rates depends on elevation changes, and not on the size of grains or the length of the control volume.
Expanding the entire first and second terms to second order $z_1$ provides the Fokker-Planck equation
\be 0 = -2\overline{\kappa}z_1[zM(z)]' + z_1^2 M''(z). \ee
Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, we find
\be M = M_0e^{-\kappa (z/z_1)^2}\ee
as stated in the manuscript.

\subsection{Closure equation approach for bed elevations}
This section produces an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
\be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
for the mean particle activity conditional to the elevation $m$ into into manuscript equation (9), noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
\begin{multline} 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]\\-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
\end{multline}
At this point, $c$ is considered an undetermined positive constant that may depend on the entrainment, migration, and deposition rates.
Taking the Fokker-Planck expansion and requiring the distribution $M(m)$ to vanish at infinity for normalization yields
\be 0 = 2\kappa m \Big( 2 + \frac{(\sigma-\mu)c}{E}\Big)M(m) + \Big\{
\Big[2-\kappa m \frac{(\sigma+\mu)c}{E} \Big]M(m)\Big\}'.\ee
Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$, we find
\be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}.\ee
Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, we determine the closure relation
\be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation (12) of the manuscript that is plotted with the numerical simulations in manuscript figure (4a).

### 2021-07-07 app_ch5.tex
progress ?

Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
\begin{align} P(u,t+\delta t) =
	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
\end{align}
Evaluating all integrals over $\delta$-functions provides 
\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}. This process finally obtains the master equation (\ref{eq:master}).
\section{Derivation of Steady-state solution}
\label{sec:langsteadyderiv}
Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
where
\be q(z) = d z^2 - i \gamma z + 1 . \label{eq:q} \ee
Recursing $N+1$ times provides
\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
\be \lambda_\pm = \frac{\gamma}{2d}\Big[1 \pm \sqrt{ 1 + 4d/\gamma^2} \Big].\ee
Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
where the coefficients $R_l^\pm$ are the residues of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
The Fourier transform (\ref{eq:recursion}) has a beautiful feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
Taking this limit and evaluating the residues provides 
\begin{multline} \tilde{P}(s) = \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
	\\+ \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
Finally, inverting the Fourier transforms term by term with contour integration and incorporating (\ref{eq:q}) provides the steady-state solution (\ref{eq:steadystate}).

\section{Calculation of the moments}
\label{sec:langmoments}
Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
\be 0 = Dk (k-1)\langle u^{k-2} \rangle + \Gamma k \langle u^{k-1} \rangle + \nu (\ve^k-1)\langle u^k \rangle. \ee
$k=1$ provides the mean
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}\ee
while $k=2$ provides the second moment
\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance
\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
\section{Weak and strong collision limits}
\label{sec:langextremes}
Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities. The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using 
\be z = \frac{u - \bar{u}}{\sigma_u} \ee
and 
\be Q(z) = \sigma_u P(u)\ee
to seek a differential equation for $Q(z)$ with manageable behavior as $\ve \rightarrow 1$.
Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" Master equation
\be (1-\ve^2) \frac{d}{2d + \gamma} Q''(z) - \frac{\gamma\sqrt{1-\ve^2}}{\sqrt{2d+\gamma^2}}Q'(z) - Q(z) + \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\gamma\sqrt{1-\ve^2}}{\ve\sqrt{2d+\gamma^2}} \Big]\Big) = 0 \ee
This equation remains exact and is only a change of variables from (\ref{eq:master}).
Now we approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$. This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

>>>>

Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
\begin{align} P(u,t+\delta t) =
	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
\end{align}
Evaluating all integrals over $\delta$-functions provides 
\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}. This process finally obtains the master equation (\ref{eq:master}).
\section{Derivation of Steady-state solution}
\label{sec:langsteadyderiv}
Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
where
\be q(z) = \tilde{D} z^2 - i \tilde{\Gamma} z + 1 . \label{eq:q} \ee
Recursing $N+1$ times provides
\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
\be \lambda_\pm = \frac{\tilde{\Gamma}}{2\tilde{D}}\Big[1 \pm \sqrt{ 1 + 4\tilde{D}/\tilde{\Gamma}^2} \Big]. \label{eq:lambdas}\ee
Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
where the coefficients $R_l^\pm$ are the residues (at least up to a constant factor) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
The Fourier transform (\ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
Taking this limit and evaluating the residues provides 
\begin{multline} \tilde{P}(s) = \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
Finally, inverting the Fourier transforms term by term with contour integration (another use of the integral given in eq. \ref{eq:contour}), and incorporating eq. (\ref{eq:q}), provides the steady-state solution (\ref{eq:steadystate}).

\section{Calculation of the moments}
\label{sec:langmoments}
Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
\be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
$k=1$ provides the mean (given in eq. \ref{eq:meanu})
\be \langle u \rangle = \frac{\tilde{\Gamma}}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}\ee
while $k=2$ provides the second moment
\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance
\be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
given in equation \ref{eq:varu}.
\section{Weak and strong collision limits}
\label{sec:langextremes}
Now I demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities. The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using the scaled variables
\be z = \frac{u - \bra u \ket}{\sigma_u}, \ee
and
\be Q(z) = \sigma_u P(u).\ee
This produces a differential equation for $Q(z)$ which is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" master equation
\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}
This equation remains exact and is only a change of variables from (\ref{eq:master}).
Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-07-07 ch5langevin.tex
progress ?

In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(-d \lambda_-^2 \ve^{-2m} + \gamma \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in the appendix; they are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors are $K_\pm$ are 
\be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution appears complicated, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.

It is possible to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
These calculations are provided in appendix section \ref{sec:langmoments}, where the mean bedload velocity is derived as
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}.\ee
This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
The second moment is
\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}.\ee
This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are a leading control on particle velocity fluctuations.

>>>>

In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors are $K_\pm$ are 
\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values. The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.

It is possible to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
These calculations are provided in appendix section \ref{sec:langmoments}, where the mean bedload velocity is derived as
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}. \label{eq:meanu}\ee
This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
The second moment is
\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}. \label{eq:varu}\ee
This result demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are a leading control on particle velocity fluctuations.
